<<<<<<< HEAD

# Alina part - Numeric

```{r dataload}
library(readxl)
Data <- read_excel("C:/Users/Bruger/OneDrive - Aarhus universitet/8. semester - BI/ML2 - Machine Learning 2/ML2EXAM/Data.xls")
empNumeric <- Data
```

```{r}
#install.packages("visdat")
```


```{r packages}
# Helper packages
library(dplyr)    # for data manipulation
library(ggplot2)  # for awesome graphics
library(visdat)   # for additional visualizations
# Feature engineering packages
library(caret)    # for various ML tasks
library(recipes)  # for feature engineering tasks
```


```{r data structure - overview}
library(dplyr)
# Convert all character columns in the data frame 'emp' to factors
empNumeric <- empNumeric %>%
  mutate(across(where(is.character), factor))
glimpse(empNumeric)
```
character string (<chr>): these should be transformed into categorical variables
numeric <dbl>

```{r Missing overview}
library(visdat)
sum(is.na(empNumeric))
vis_miss(empNumeric, cluster = TRUE) #visdat library
plot_missing(empNumeric)
```
- no NA's in variables: reviewId, reviewDateTime, ratingOverall, ratingWorkLifeBalance, ratingCultureAndValues, -	ratingDiversityAndInclusion, ratingSeniorLeadership, ratingCareerOpportunities, ratingCompensationAndBenefits, lengthOfEmployment
- NA's from 0% - 5%: employmentStatus
- NA's from 5% - 10%: jobTitle.text
- NA's from 10% - 40%: ratingRecommendToFriend, isCurrentJob, location.name
- NA's from 40% -> : jobEndingYear


## reviewId: Unique identifier for each review

```{r reviewId}
library(tidyverse)
library(DataExplorer)
str(empNumeric$reviewId)
empNumeric$reviewId <- NULL
```
DELETE THIS VARIABLE

## reviewDateTime: Timestamp of when the review was submitted

```{r reviewDateTime }
str(empNumeric$reviewDateTime)
```
follows the standard ISO 8601 format: YYYY-MM-DDTHH:MM:SS.fff, where:
YYYY represents the year,
MM the month,
DD the day,
T is a separator (indicating the start of the time portion),
HH the hour (in 24-hour format),
MM the minutes,
SS the seconds,
fff the milliseconds.
- given this interpretation, the information after the T-separator is evaluated as being redundant. Therefore the variable is being transformed into a numeric obtaining the Year variable

```{r Create year}
library(lubridate) 
# Convert 'reviewDateTime' from character to POSIXct format (if not already)
empNumeric$reviewDateTime <- ymd_hms(empNumeric$reviewDateTime)

# Extract the year and convert it to numeric format
empNumeric$reviewYear <- year(empNumeric$reviewDateTime)
```
reviewYear is created being a numeric with range 2013->2024.

```{r delete origin variable}
empNumeric$reviewDateTime <- NULL 
```

```{r EDA reviewYear}
library(tidyverse)
library(DataExplorer)
library(ggplot2)

# Create a bar plot
ggplot(empNumeric, aes(x = as.factor(reviewYear))) +
  geom_bar(stat = "count", fill = "blue", color = "black") +  # Count is default, explicitly stating for clarity
  labs(x = "Review Year", y = "Number of Reviews", title = "Distribution of Reviews by Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x labels for better visibility if needed

summary(empNumeric$reviewYear)

```
The reviewDateTime variable has been split into
reviewYear
- a numeric variable having the vast majority of observations from the year 2023. The next coming years 2022 and 2024 do also represent a fair amount of the observations. The years including 2021 and before, are bearing a very minor part of the years, as these reviews could be consideres deleted.


## ratingOverall: Overall rating given to the company by the employee
This is the variable to predict based on the other features

```{r Factorize/ordering ratingOverall}
#empNumeric$ratingOverall <- factor(empNumeric$ratingOverall)
summary(empNumeric$ratingOverall)
# Convert to an ordinal factor
#empNumeric$ratingOverall <- factor(empNumeric$ratingOverall, levels = c(1, 2, 3, 4, 5), ordered = TRUE)
```
The average perception of the company is a rating of 3.76 from all employees


```{r Histogram ratingOverall }
histogram(empNumeric$ratingOverall)

# Load the ggplot2 library
library(ggplot2)

# Create a histogram of ratingOverall
ggplot(empNumeric, aes(x = ratingOverall)) +
  geom_histogram(stat = "count", fill = "steelblue", color = "black", binwidth = 1) +
  labs(x = "Overall Rating", y = "Frequency", title = "Distribution of Overall Ratings") +
  theme_minimal() +
  scale_x_discrete(limits = c("1", "2", "3", "4", "5"))  # Ensure that all rating levels are shown even if some have 0 counts
```
Discuss wether ratingOverall yields the best insights being numeric or being a factor?
Do we want to obtain RMSE measures or Accuracy measures? 


#DEALING WITH MISSING

```{r isCurrentJob replacing NAs with 0}
library(dplyr)
# Replace NA values with 0s
empNumeric <- empNumeric %>% 
  mutate(isCurrentJob = replace_na(isCurrentJob, 0))
summary(empNumeric$isCurrentJob)
```
isCurrentJob:
- the 39% missing values are being evaluated as "Informative missing values" (Kuhn and Johnson 2013). As the variable only contains 1s and NAs, these NAs are indeed an informative missing value which requires being transformed into 0s.


Deleting variables: jobEndingYear,jobTitle.text,location.name

jobEndingYear
- deleting this variable due to the 61% Missings. As support for this deletion, the variable isCurrentJob is evaluated as having an adequate extent of information given in jobEndingYear.

jobTitle.text
- Deleting this variable as we have a huge amount of different jobtitles, namely 2733 various instances, whereas these do not contribute with anything but noise to the dataset.

location.name
- Deleting this variable as we have a huge amount of different location anmes, namely 1201 various instances, whereas these do not contribute with anything but noise to the dataset.


```{r Delete jobEndingYear;jobTitle.text;location.name}
empNumeric <- subset(empNumeric, select = -c(jobEndingYear))
empNumeric <- subset(empNumeric, select = -c(jobTitle.text))
empNumeric <- subset(empNumeric, select = -c(location.name))
```

Deleting NAs for the four variables and prevent making imputations, because it is assumed that it is easier to impute a value there is numeric compared to a variable there are based on your feelings and opinions
```{r Delete NAs ratingCeo;ratingBusinessOutlook;employmentStatus;ratingRecommendToFriend}
empNumeric <- empNumeric[!is.na(empNumeric$ratingCeo), ]
empNumeric <- empNumeric[!is.na(empNumeric$ratingBusinessOutlook), ]
empNumeric <- empNumeric[!is.na(empNumeric$employmentStatus), ]
empNumeric <- empNumeric[!is.na(empNumeric$ratingRecommendToFriend), ]

sum(is.na(empNumeric))
```

From Exploratory Data Analysis is seen in the dataset, that 310 respondents is scoring ratingOverall with a value from 3-5 and at the same score 
- ratingCareerOpportunities,
- ratingCompensationAndBenefits,
- ratingCultureAndValues,
- ratingDiversityAndInclusion,
- ratingSeniorLeadership,
- ratingWorkLifeBalance
to zero. This is assessed being a flawed respondent, as it is assumed that this respondent does not take a stand on the above listed question-categories.
```{r deleting instances having 0s in several categorical 0-5 and ratingOverall 3-5}
# Since there's still some giving 0's another subset of data will be made, deleting all the rows where 
# the respondant answered 0
empNumeric <- empNumeric[!(empNumeric$ratingCareerOpportunities == 0 |
                         empNumeric$ratingCompensationAndBenefits == 0 |
                         empNumeric$ratingCultureAndValues == 0 |
                         empNumeric$ratingDiversityAndInclusion == 0 |
                         empNumeric$ratingSeniorLeadership == 0 |
                         empNumeric$ratingWorkLifeBalance == 0), ]
plot_histogram(empNumeric)
```

ratingCeo, ratingBusinessOutlook, ratingRecommendToFriend, employmentStatus
- transforming these variables into numeric variables.

Assumptions made in this step:
- ratingCeo converted "APPROVE" to 2, "NO_OPINION" to 1, and "DISAPPROVE" to 0. Here is made a ranking scale of the three categories.

- ratingBusinessOutlook converted "POSITIVE" to 2, "NEUTRAL" to 1, and "NEGATIVE" to 0. Here is made a ranking scale of the three categories.

- ratingRecommendToFriend converted "POSITIVE" to 1 and "NEGATIVE" to 0. Here is made a ranking scale of the two categories.

- employmentStatus converted "REGULAR" to 1 and "PART_TIME" to 0. Here is made a ranking scale of the two categories.


```{r numeric transformation variables }
#TRANSFORMING INTO NUMERIC SCALE

###ratingCeo
# Convert "APPROVE" to 2, "NO_OPINION" to 1, and "DISAPPROVE" to 0
empNumeric$ratingCeo <- factor(empNumeric$ratingCeo, levels = c("DISAPPROVE", "NO_OPINION", "APPROVE"), labels = c(0, 1, 2))

# Convert the factor with numeric labels to numeric type
empNumeric$ratingCeo <- as.numeric(as.character(empNumeric$ratingCeo))

###ratingBusinessOutlook
# Convert "POSITIVE" to 2, "NEUTRAL" to 1, and "NEGATIVE" to 0
empNumeric$ratingBusinessOutlook <- factor(empNumeric$ratingBusinessOutlook, levels = c("NEGATIVE", "NEUTRAL", "POSITIVE"), labels = c(0, 1, 2))

# Convert the factor with numeric labels to numeric type
empNumeric$ratingBusinessOutlook <- as.numeric(as.character(empNumeric$ratingBusinessOutlook))

###ratingRecommendToFriend
# Convert "POSITIVE" to 1 and "NEGATIVE" to 0
empNumeric$ratingRecommendToFriend <- ifelse(empNumeric$ratingRecommendToFriend == "POSITIVE", 1, 0)

empNumeric$ratingRecommendToFriend <- as.numeric(empNumeric$ratingRecommendToFriend)

###employmentStatus
empNumeric$employmentStatus <- ifelse(empNumeric$employmentStatus == "REGULAR", 1, 0)

empNumeric$employmentStatus <- as.numeric(empNumeric$employmentStatus)

glimpse(empNumeric)
```

# Data Preparation

```{r datasplit standard}
library(rsample)
set.seed(123) # Set a random seed for replication purposes
numericsplit <- initial_split(empNumeric, prop = 0.70, strata = "ratingOverall")
numerictrain  <- training(numericsplit)
numerictest   <- testing(numericsplit)
dim(numerictrain)
dim(numerictest)
```

```{r Recipe standard }
#Creating the blueprint
numericrecipe <- recipe(ratingOverall ~ ., data = empNumeric) %>%
  step_normalize(all_numeric(), -all_outcomes()) %>%
  #step_impute_knn(all_predictors(), neighbors = 5) %>%
  #step_BoxCox(all_outcomes()) %>%
  #step_YeoJohnson(all_numeric(), -all_outcomes()) %>%
  # step_other(all_nominal(), threshold = 0.05, other = "Other") %>%  #lumping if needed
  step_center(all_numeric(), -all_outcomes()) %>%
  step_scale(all_numeric(), -all_outcomes()) %>%
  step_dummy(all_nominal(), -all_outcomes(), one_hot = FALSE) %>% #produces dummy encoding (True is one-hot encoding)
  step_nzv(all_predictors()) # using NZV last removes X factor levels with near zero variance
numericrecipe

#Preparing the blueprint based on training data
numericprepare <- prep(numericrecipe, training = numerictrain)
numericprepare

#baking: applying the recipe to new data / test data
numericbaked_train <- bake(numericprepare, new_data = numerictrain)
numericbaked_test <- bake(numericprepare, new_data = numerictest)
```

```{r dim standard numericrecipe}
dim(numericbaked_train)
dim(numericbaked_test)
```

If computation is heavy
```{r downsizing datasplit }
set.seed(123)
random_indices <- sample(1:nrow(empNumeric), 1000)
sub_empNumeric<- empNumeric[random_indices, ]

subsplitnumeric  <- initial_split(sub_empNumeric, prop = 0.7, strata = "ratingOverall")
subtrainnumeric <- training(subsplitnumeric)
subtestnumeric  <- testing(subsplitnumeric)
```

```{r Recipe subset }
#Creating the blueprint
subrecipenumeric <- recipe(ratingOverall ~ ., data = sub_empNumeric) %>%
  step_normalize(all_numeric(), -all_outcomes()) %>%
  #step_impute_knn(all_predictors(), neighbors = 5) %>%
  #step_BoxCox(all_outcomes()) %>%
  #step_YeoJohnson(all_numeric(), -all_outcomes()) %>%
  # step_other(all_nominal(), threshold = 0.05, other = "Other") %>%  #lumping if needed
  step_center(all_numeric(), -all_outcomes()) %>%
  step_scale(all_numeric(), -all_outcomes()) %>%
  step_dummy(all_nominal(), -all_outcomes(), one_hot = FALSE) %>% #produces dummy encoding (True is one-hot encoding)
  step_nzv(all_predictors()) # using NZV last removes X factor levels with near zero variance
subrecipenumeric

#Preparing the blueprint based on training data
subpreparenumeric <- prep(subrecipenumeric, training = subtrainnumeric)
subpreparenumeric

#baking: applying the recipe to new data / test data
subbaked_trainnumeric <- bake(subpreparenumeric, new_data = subtrainnumeric)
subbaked_testnumeric <- bake(subpreparenumeric, new_data = subtestnumeric)
```


```{r dimensions baked train/test}
#New datasets
dim(subbaked_trainnumeric)
dim(subbaked_testnumeric)
```

##8.4.8 EXE 8 Regression tree and bagging

Regression tree and bagging - REGRESSION PROBLEM
- In the lab, a classification tree was applied to the Carseats data set after converting Sales into a qualitative response variable. Now we will seek to predict Sales using regression trees and related approaches, treating the response as a quantitative variable.
- (a) Split the data set into a training set and a test set.
- (b) Fit a regression tree to the training set. Plot the tree, and interpret the results. What test MSE do you obtain?
- (c) Use cross-validation in order to determine the optimal level of tree complexity. Does pruning the tree improve the test MSE?
- (d) Use the bagging approach in order to analyze this data. What test MSE do you obtain? Use the importance() function to determine which variables are most important.
- (e) Use random forests to analyze this data. What test MSE do you obtain? Use the importance() function to determine which variables are most important. Describe the effect of m, the number of variables considered at each split, on the error rate obtained.
- (f) Now analyze the data using BART, and report your results.

- (a) Split the data set into a training set and a test set.
```{r datasplit subbaked_trainnumeric + test}
dim(numerictrain)
dim(numerictest)
```

- (b) Fit a regression tree to the training set. Plot the tree, and interpret the results. What test MSE do you obtain?
```{r Fit tree}
library(tree)
tree.carseats = tree(ratingOverall ~ ., numerictrain)
summary(tree.carseats)
# Notice in the summary output that variables actually used in tree construction
# Note also that this (unpruned) tree has 19 terminal nodes
```
Regression tree:
Variables actually used in tree construction:
[1] "ratingCultureAndValues"    "ratingCareerOpportunities"
[3] "ratingSeniorLeadership"    "ratingBusinessOutlook"

Number of terminal nodes:  7
- Terminal nodes, or leaf nodes, are the end points of the tree where predictions are made. 
- The number of terminal nodes can give you an idea of the complexity of the model: more nodes generally mean a more complex model.

Residual mean deviance:  0.5732 = 3250 / 5671
- deviance is simply the sum of squared errors for the tree
-  calculated as the total deviance divided by the degrees of freedom (number of observations minus the number of terminal nodes).
- Lower deviance values generally suggest a better fit of the model to the data.

Distribution of residuals:
  Min.     1st Qu.  Median    Mean 3rd Qu.    Max. 
-3.7400    -0.4817  0.2109  0.0000  0.2602  3.5180
- describing the behavior of residuals in the model

use the plot() function to display the tree structure, and the text() function to display the node labels. The argument pretty = 0 instructs R to include the category names for any qualitative predictors, rather than simply displaying a letter for each category
```{r Tree plot}
# Now plot it
plot(tree.carseats)
text(tree.carseats, pretty = 0)
```
- The root of the tree splits the data based on ratingCultureAndValues < 2.5. This suggests that ratingCultureAndValues is a significant predictor in determining the outcome.
- Deeper in the tree, other variables are presented with decisions


```{r Tree metrics and measures}
tree.carseats
```
1) root 5678 8691.0 3.760  
5678: all observations in df
8691.0: deviance
3.760: predicted mean value of the target variable for this node (3.760). This is the average outcome for all observations at the root before any splits are made.

2) ratingCultureAndValues < 2.5 1326 1654.0 2.275  
2): first split
ratingCultureAndValues < 2.5: the tree first splits the data based on whether this rating is above/below 2,5
1326: 
1654.0: deviance for each node, providing a measure of the tree's error post-split.
2.275: mean value of the target variable for observations in each node.

We extract this above table to gain accurate information about the decision tree nodes


Evaluating performance of a classification tree on these data, the test error is estimated. Splitting the observations into a training set and a test set, building the tree using the training set, and evaluate its performance on the test data. 
```{r Tree test predictions}
pred.trees <- predict(tree.carseats, numerictest)
mean((numerictest$ratingOverall - pred.trees)^2)
```
the test MSE is about 0.6013

### 8.4.8 EXE 8 - CV + Pruning

- (c) Use cross-validation in order to determine the optimal level of tree complexity. Does pruning the tree improve the test MSE?

cv.tree() runs a K-fold cross-validation experiment to find the deviance or number of misclassifications as a function of the cost-complexity parameter k (alpha, in the textbook).
```{r CV choosing level of trees}
set.seed(123)
cv.carseats = cv.tree(tree.carseats, FUN = prune.tree) 
par(mfrow = c(1, 2))
plot(cv.carseats$size, cv.carseats$dev, type = "b")
plot(cv.carseats$k, cv.carseats$dev, type = "b")

cv.carseats$dev 
min(cv.carseats$dev)
# The minimum cv-error is obtained for the size=7 OBS for reversed dev in this list!!??
```
plotting the MSE as a function of the number of leaves $size (number of terminal nodes, labelled, T, see p. 309 in textbook)
- discussing the number of terminal nodes as a trade-off from deviance/MSE

Complexity Parameter (k): 
- the plot of k vs. deviance helps in understanding how sensitive the tree is to the pruning process. A sharp increase in deviance as k increases indicates a point beyond which the tree loses significant predictive accuracy

```{r Pruned Tree plot}
# We can now prune the original tree: 
  # Best size = 7 - BUT WE CHOOSE 5 as getting something comparable
pruned.carseats = prune.tree(tree.carseats, best = 5)
par(mfrow = c(1, 1))
plot(pruned.carseats)
text(pruned.carseats, pretty = 0)
```

Evaluating performance of a classification tree on these data, the test error is estimated. Splitting the observations into a training set and a test set, building the tree using the training set, and evaluate its performance on the test data. 
```{r Tree test predictions}
pred.pruned <- predict(pruned.carseats, numerictest)
mean((numerictest$ratingOverall - pred.pruned)^2)
```
the test MSE is 0.6252, which is 0.0239 better than the standard tree. (0.6252-0.6013 = 0.0239) 

### 8.4.8 EXE 8 Bagging

- (d) Use the bagging approach in order to analyze this data. What test MSE do you obtain? Use the importance() function to determine which variables are most important.

```{r}
library(randomForest)
set.seed(123)
bag.carseats = randomForest(ratingOverall ~ ., data = numerictrain, mtry = 10, ntree = 500, importance = T) # notice mtry = p

bag.pred = predict(bag.carseats, Carseats.test)
mean((Carseats.test$Sales - bag.pred)^2)  # Bagging improves the test MSE to approx 2.5

# The relative importance of the variables: 
importance(bag.carseats)
# We see that Price, ShelveLoc, CompPrice, Adv 
# and Age are the most important variables to predict sales ï¿½ðšŽ
varImpPlot(bag.carseats)
```
HERE IS STATES THAT RF does not work on numeric DV's^^^^


- (e) Use random forests to analyze this data. What test MSE do you obtain? 
Use the importance() function to determine which variables are most important. Describe the effect of m, the number of variables considered at each split, on the error rate obtained.
```{r}
library(randomForest)
set.seed(123)
rf.carseats = randomForest(ratingOverall ~ ., data = numerictrain, mtry = 5, ntree = 50, importance = T) 
## notice mtry = < p; 
## Changing mtry (m) that is, the number of variables considered at each split,varies test MSE.

rf.pred = predict(rf.carseats, Carseats.test)
mean((Carseats.test$Sales - rf.pred)^2) # Random forest improves even more the MSE on test set to 2.28
importance(rf.carseats)
# We see again the most important variables to predict sales
varImpPlot(rf.carseats)
```
HERE IS STATES THAT RF does not work on numeric DV's^^^^























# from Ralle + Boysen MODEL DEVELOPMENT

```{r Boysen}
library(rsample)
set.seed(123) # Set a random seed for replication purposes
job_split <- initial_split(emp, prop = 0.80, strata = "ratingOverall")
job_train  <- training(job_split)
job_test   <- testing(job_split)
=======

# Alina part - Numeric

```{r dataload}
library(readxl)
Data <- read_excel("C:/Users/Bruger/OneDrive - Aarhus universitet/8. semester - BI/ML2 - Machine Learning 2/ML2EXAM/Data.xls")
empNumeric <- Data
```

```{r}
#install.packages("visdat")
```


```{r packages}
# Helper packages
library(dplyr)    # for data manipulation
library(ggplot2)  # for awesome graphics
library(visdat)   # for additional visualizations
# Feature engineering packages
library(caret)    # for various ML tasks
library(recipes)  # for feature engineering tasks
```


```{r data structure - overview}
library(dplyr)
glimpse(empNumeric)
```
character string (<chr>): these should be transformed into categorical variables
numeric <dbl>

```{r Missing overview}
library(visdat)
sum(is.na(empNumeric))
vis_miss(empNumeric, cluster = TRUE) #visdat library
plot_missing(empNumeric)
```
- no NA's in variables: reviewId, reviewDateTime, ratingOverall, ratingWorkLifeBalance, ratingCultureAndValues, -	ratingDiversityAndInclusion, ratingSeniorLeadership, ratingCareerOpportunities, ratingCompensationAndBenefits, lengthOfEmployment
- NA's from 0% - 5%: employmentStatus
- NA's from 5% - 10%: jobTitle.text
- NA's from 10% - 40%: ratingRecommendToFriend, isCurrentJob, location.name
- NA's from 40% -> : jobEndingYear


## reviewId: Unique identifier for each review

```{r reviewId}
library(tidyverse)
library(DataExplorer)
str(empNumeric$reviewId)
empNumeric$reviewId <- NULL
```
DELETE THIS VARIABLE

## reviewDateTime: Timestamp of when the review was submitted

```{r reviewDateTime }
str(empNumeric$reviewDateTime)
```
follows the standard ISO 8601 format: YYYY-MM-DDTHH:MM:SS.fff, where:
YYYY represents the year,
MM the month,
DD the day,
T is a separator (indicating the start of the time portion),
HH the hour (in 24-hour format),
MM the minutes,
SS the seconds,
fff the milliseconds.
- given this interpretation, the information after the T-separator is evaluated as being redundant. Therefore the variable is being transformed into a numeric obtaining the Year variable

```{r Create year}
library(lubridate) 
# Convert 'reviewDateTime' from character to POSIXct format (if not already)
empNumeric$reviewDateTime <- ymd_hms(empNumeric$reviewDateTime)

# Extract the year and convert it to numeric format
empNumeric$reviewYear <- year(empNumeric$reviewDateTime)
```
reviewYear is created being a numeric with range 2013->2024.

```{r EDA reviewYear}
library(tidyverse)
library(DataExplorer)
library(ggplot2)

# Create a bar plot
ggplot(empNumeric, aes(x = as.factor(reviewYear))) +
  geom_bar(stat = "count", fill = "blue", color = "black") +  # Count is default, explicitly stating for clarity
  labs(x = "Review Year", y = "Number of Reviews", title = "Distribution of Reviews by Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x labels for better visibility if needed

summary(empNumeric$reviewYear)

```
The reviewDateTime variable has been split into
reviewYear
- a numeric variable having the vast majority of observations from the year 2023. The next coming years 2022 and 2024 do also represent a fair amount of the observations. The years including 2021 and before, are bearing a very minor part of the years, as these reviews could be consideres deleted.


## ratingOverall: Overall rating given to the company by the employee
This is the variable to predict based on the other features

```{r Factorize/ordering ratingOverall}
empNumeric$ratingOverall <- factor(empNumeric$ratingOverall)
summary(empNumeric$ratingOverall)
# Convert to an ordinal factor
empNumeric$ratingOverall <- factor(empNumeric$ratingOverall, levels = c(1, 2, 3, 4, 5), ordered = TRUE)
```
Making the variable ordinal scaled: ordered = TRUE

```{r Histogram ratingOverall }
histogram(empNumeric$ratingOverall)

# Load the ggplot2 library
library(ggplot2)

# Create a histogram of ratingOverall
ggplot(empNumeric, aes(x = ratingOverall)) +
  geom_histogram(stat = "count", fill = "steelblue", color = "black", binwidth = 1) +
  labs(x = "Overall Rating", y = "Frequency", title = "Distribution of Overall Ratings") +
  theme_minimal() +
  scale_x_discrete(limits = c("1", "2", "3", "4", "5"))  # Ensure that all rating levels are shown even if some have 0 counts
```
Discuss wether ratingOverall yields the best insights being numeric or being a factor?
Do we want to obtain RMSE measures or Accuracy measures? 

COMMON ANSWER: make into ordinal scale

#DEALING WITH MISSING

```{r isCurrentJob replacing NAs with 0}
library(dplyr)
# Replace NA values with 0s
empNumeric <- empNumeric %>% 
  mutate(isCurrentJob = replace_na(isCurrentJob, 0))
summary(empNumeric$isCurrentJob)
```
isCurrentJob:
- the 39% missing values are being evaluated as "Informative missing values" (Kuhn and Johnson 2013). As the variable only contains 1s and NAs, these NAs are indeed an informative missing value which requires being transformed into 0s.


Deleting variables: jobEndingYear,jobTitle.text,location.name

jobEndingYear
- deleting this variable due to the 61% Missings. As support for this deletion, the variable isCurrentJob is evaluated as having an adequate extent of information given in jobEndingYear.

jobTitle.text
- Deleting this variable as we have a huge amount of different jobtitles, namely 2733 various instances, whereas these do not contribute with anything but noise to the dataset.

location.name
- Deleting this variable as we have a huge amount of different location anmes, namely 1201 various instances, whereas these do not contribute with anything but noise to the dataset.


```{r Delete jobEndingYear;jobTitle.text;location.name}
empNumeric <- subset(empNumeric, select = -c(jobEndingYear))
empNumeric <- subset(empNumeric, select = -c(jobTitle.text))
empNumeric <- subset(empNumeric, select = -c(location.name))
```

Deleting NAs for the four variables and prevent making imputations, because it is assumed that it is easier to impute a value there is numeric compared to a variable there are based on your feelings and opinions
```{r Delete NAs ratingCeo;ratingBusinessOutlook;employmentStatus;ratingRecommendToFriend}
empNumeric <- empNumeric[!is.na(empNumeric$ratingCeo), ]
empNumeric <- empNumeric[!is.na(empNumeric$ratingBusinessOutlook), ]
empNumeric <- empNumeric[!is.na(empNumeric$employmentStatus), ]
empNumeric <- empNumeric[!is.na(empNumeric$ratingRecommendToFriend), ]

sum(is.na(empNumeric))
```

From Exploratory Data Analysis is seen in the dataset, that 310 respondents is scoring ratingOverall with a value from 3-5 and at the same score 
- ratingCareerOpportunities,
- ratingCompensationAndBenefits,
- ratingCultureAndValues,
- ratingDiversityAndInclusion,
- ratingSeniorLeadership,
- ratingWorkLifeBalance
to zero. This is assessed being a flawed respondent, as it is assumed that this respondent does not take a stand on the above listed question-categories.
```{r deleting instances having 0s in several categorical 0-5 and ratingOverall 3-5}
# Since there's still some giving 0's another subset of data will be made, deleting all the rows where 
# the respondant answered 0
empNumeric <- empNumeric[!(empNumeric$ratingCareerOpportunities == 0 |
                         empNumeric$ratingCompensationAndBenefits == 0 |
                         empNumeric$ratingCultureAndValues == 0 |
                         empNumeric$ratingDiversityAndInclusion == 0 |
                         empNumeric$ratingSeniorLeadership == 0 |
                         empNumeric$ratingWorkLifeBalance == 0), ]
plot_histogram(empNumeric)
```

ratingCeo, ratingBusinessOutlook, ratingRecommendToFriend, employmentStatus
- factorizing these variables into categorical variables.

lengthOfEmployment
- scaling lenght of employment
```{r Factorize categorical variables + scaling}
empNumeric$ratingCeo <- factor(empNumeric$ratingCeo)
empNumeric$ratingBusinessOutlook <- factor(empNumeric$ratingBusinessOutlook)
empNumeric$ratingRecommendToFriend <- factor(empNumeric$ratingRecommendToFriend)
empNumeric$employmentStatus <- factor(empNumeric$employmentStatus)

empNumeric$lengthOfEmployment <- scale(empNumeric$lengthOfEmployment)

glimpse(empNumeric)
```

# MODEL DEVELOPMENT

```{r Boysen}
library(rsample)
set.seed(123) # Set a random seed for replication purposes
job_split <- initial_split(emp, prop = 0.80, strata = "ratingOverall")
job_train  <- training(job_split)
job_test   <- testing(job_split)
>>>>>>> 63dc1c65cf50349f1b453a9fccae7d737922855f
```